---
title: "GLMM GRAN Vegetation"
author: "Marte Fandrem"
format: pdf
editor: visual
---

```{r Packages needed, include=FALSE, return=FALSE, warning=FALSE}
library(tidyverse)
library(lme4) 
library(merTools) 
library(glmmTMB)
library(brms)
library(modelr)
library(nlme)

```

### Upload and clean dataset

```{r Upload datasets, include=FALSE}
df <- readr::read_delim('C:/Users/martef/DokumenterIntern/GitHub/GRAN_Revegetation/Data/Revegetering_alldata.csv', 
                                  delim = ';',
                                  locale = locale('se', encoding = 'ISO8859-1'),
                                  col_names = FALSE)

#need to switch rows and columns
reveg = data.frame(t(df)) #Transposing
names(reveg) <- reveg[1,] #Copying header names from row 1
reveg <- reveg[-1,] #Removing row 1

reveg$year <- substr(reveg$Date, 1,4) #Creating columns for year and month
reveg$month <- substr(reveg$Date, 6,7)
reveg <- reveg%>% rename(Location = Localition) #Fix bad spelling
reveg[reveg=='<1'] <- '0.1' #change all <1 to 0.1
reveg <- reveg %>% mutate_at(c('Location', 'Sublocation', 'Block', 'Treatment', 'year', 'month'), as.factor)
reveg <- reveg %>% mutate_at(c(7:36), as.numeric) #Change from character to numeric on multiple columns
reveg <- reveg %>% 
  mutate_at(c(9:36), ~replace_na(.,0)) #Replacing NAs with zeros

reveg_long <-reveg%>%
  pivot_longer(Sphagnum:Temp_water, names_to = 'functional_group', values_to = "cover") 

```

**My study questions:**

1.  How well does Sphagnum establish when applied as fragments onto bare peat compared to the controls?
2.  Will altering the microenvironment by adding mulch alone give a positive effect on natural establishment of Sphagnum and other plants compared to the controls?
3.  Is the establishment rate of Sphagnum cover related to species specific traits of the Sphagnum species?
4.  How well does the treatments compare to reference sites relative to the controls?

**I'll start with analysing Sphagnum cover.**

## Sphagnum cover

I want to investigate the relationship between Sphagnum cover and time since establishment compared with controls and mulch treatment.

This is tested with generalized linear mixed models, as my values are nested:

*cover **value** in percentage within a treatment (0-100%)\<- **block** of treatments (C, M, Sx2) \<- repeated treatments on a **sublocation** (4 blocks)\<- repeated sublocations on a **location** (3 (or 2))*

y\~ax+b
